---
title: "DSI Symposium Workshop"
author: "Mateo Aristizabal"
date: "2024-03-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installing the Packages

```{r include=FALSE}
install.packages("rpart")
install.packages("caret", dependencies = TRUE)
install.packages("dplyr")
install.packages("rpart.plot")
install.packages("data.tree")
install.packages("caTools")

```

## Loading the Packages

```{r}
library(rpart)
library(caret)
library(rpart.plot)
library(dplyr)
library(data.tree)
library(caTools)
library(caTools)
library(ggplot2)
library(pROC)
```

# Initializing the Dataset
```{r}
data <- read.csv("https://raw.githubusercontent.com/Mateo486/Decision-Trees-Workshop/main/heart[2].csv")

data
```
# Cleaning up Data Names
```{r}
data <- data %>%
  rename(
  BloodPressure = RestingBP,
  BloodSugar = FastingBS,
  ElectroCardio = RestingECG,
  MaxHeartRate = MaxHR
  )

data
```

# Recoding factor values
```{r}
data <- data %>% mutate(ChestPainTypeNew = as.factor(recode(ChestPainType,
"TA" = "Typical Angina",
"ATA" = "Atypical Angina",
"NAP" = "Non-Anginal Pain",
"ASY" = "Asymptomatic"
)))

data <- data %>% mutate(ElectroCardioNew =
as.factor(recode(ElectroCardio,
"Normal" = "Normal",
"ST" = "ST-T wave abnormality",
"LVH" = "probable/definite left ventricular hypertrophy"
)))

data <- data %>% mutate(BloodSugarNew =
as.factor(recode(BloodSugar,
"1" = ">120mg/dl",
"0" = "<120mg/dl")))

data <- data %>% mutate(HeartDiseaseNew =
as.factor(recode(HeartDisease,
"0" = "N",
"1" = "Y")))

data
```
# Data Visualization
```{r}
proportions <- prop.table(table(data$HeartDiseaseNew))

# Create a data frame for plotting
bar_data <- data.frame(
  category = names(proportions),
  proportion = proportions
)

# Plot the bar chart
bar_chart <- ggplot(bar_data, aes(x = category, y = proportions, fill = category)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(proportions * 100, 1), "%")),
            position = position_stack(vjust = 0.5)) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Proportion of Heart Disease",
       x = "Category", y = "Proportion") +
  theme_minimal()

# Print the bar chart
print(bar_chart)

```

# Selecting prediction columns
```{r}
data <- select(data, Age, Sex, ChestPainTypeNew, BloodPressure, Cholesterol, BloodSugarNew, ElectroCardioNew, MaxHeartRate, ExerciseAngina, HeartDiseaseNew)

data <- mutate(data, Sex = as.factor(Sex), ExerciseAngina = as.factor(ExerciseAngina))

data
```

#Determine Training and Testing Data: Stratified Sampling
```{r}

set.seed(12345)
index <- createDataPartition(data$HeartDiseaseNew, p = 0.8, list = FALSE, times = 1)

# Create training and testing datasets using the partition indices
train <- data[index, ]
test <- data[-index, ]
```

# Training the model
```{r}
tree <- rpart(HeartDiseaseNew ~., data = train)
```

#Predicting outcomes
```{r}
HeartAttackPredicted <- predict(tree, test, type = 'class')
```

#Confusion Matrix for evaluation
```{r}
conf_matrix <- confusionMatrix(HeartAttackPredicted, test$HeartDiseaseNew)

conf_matrix
```

#Displaying the Decision Tree
```{r}
prp(tree, varlen = 30, cex = 0.8)
```

# Classification Metrics
```{r}
accuracy <- conf_matrix$overall['Accuracy']
precision <- conf_matrix$byClass['Precision']
recall <- conf_matrix$byClass['Recall']
f1 <- conf_matrix$byClass['F1']

accuracy
precision
recall
f1

metrics_df <- data.frame(
  Metric = c('Accuracy', 'Precision', 'Recall', 'F1'),
  Value = c(accuracy, precision, recall, f1)
)

ggplot(metrics_df, aes(x = Metric, y = Value)) +
  geom_bar(stat = 'identity', fill = 'steelblue') +
  labs(title = 'Classification Metrics',
       y = 'Value') +
  theme_minimal()

plot_data <- data.frame(Age = test$Age,
                        ChestPainTypeNew = test$ChestPainTypeNew,
                        HeartDiseaseNew = test$HeartDiseaseNew)

plot_data$Color <- ifelse(plot_data$HeartDiseaseNew == "Y", "green", "red")

ggplot(plot_data, aes(x = ChestPainTypeNew, y = Age, color = Color)) +
  geom_point() +
  labs(title = "Heart Disease by Age and Chest Pain Type",
       x = "Chest Pain Type",
       y = "Age") +
  scale_color_identity() +
  theme_minimal()

```

# Receiving Operations Characteristic (ROC) Curve
```{r}


predicted_probabilities <- predict(tree, test, type = 'prob')

roc_curve <- roc(test$HeartDiseaseNew, predicted_probabilities[, "Y"])


plot(roc_curve, main = "ROC Curve for Stroke Prediction Model",
     xlab = "False Positive Rate", ylab = "True Positive Rate",
     col = "blue", lwd = 3, xaxt = "n")

abline(a = 0, b = 1, lty = 2, col = "red")

legend("bottomright", legend = c("ROC Curve", "Random Classifier"),
       col = c("blue", "red"), lty = c(1, 2), lwd = 2)


```

